---
title: "The Doppelganger Effect in Biomedical Data sets"
output: rmarkdown::github_document
author: "Wang Li Rong"
---

# 3. Exploring the ALL (Acute Lymphocytic Leukemia) Datasets

## 0) Import packages

```{r Import & Install Packages, include=FALSE}
if (!("doppelgangerIdentifier" %in% installed.packages())){
  install.packages('devtools')
  library(devtools)
  install_github('lr98769/doppelgangerIdentifier')
}
library("doppelgangerIdentifier")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!("biomaRt" %in% installed.packages())){
  BiocManager::install("biomaRt")
}
library("biomaRt")
```

File paths
```{r}
data_dir = "../data"
cleaned_data_dir = "../cleaned_data"
images_dir = "../images"
planning_data_dir = "../planning_data"
experiment_plans_dir = "../experiment_plans"
```

Load functions for pre-processing and analysis:
```{r}
functions_dir = "../functions"
source(file.path(functions_dir,"preprocessing_functions.R"))
source(file.path(functions_dir,"analysis_functions.R"))
source(file.path(functions_dir,"planning_functions.R"))
source(file.path(functions_dir,"process_vectors.R"))
source(file.path(functions_dir,"process_lists.R"))
```

## 1) Importing the Datsets
AllenData/ Ross et al.
- 15 BCR
- 18 E2A

```{r}
all_r = getDataFile(
  filename = file.path(data_dir, "Subtype-AllenData.csv"),
  affy_attribute = "affy_hg_u133a",
  batch_name = "R")
all_r_meta = getMetaDataDataframe(df = all_r,
                                  batch_name = "R")
```

MaryData/Yeoh et al.
- 15 BCR
- 27 E2A

```{r}
all_y = getDataFile(
  filename = file.path(data_dir, "Subtype-MaryData.csv"),
  affy_attribute = "affy_hg_u95av2",
  batch_name = "Y")
all_y_meta = getMetaDataDataframe(df = all_y,
                                  batch_name = "Y")
```

## 2) Finding Doppelgangers

### a) Within Ross et Al.

```{r}
doppel_all_r = getPPCCDoppelgangers(all_r, 
                                    all_r_meta,
                                    do_min_max = TRUE)
```

```{r}
visualisePPCCDoppelgangers(doppel_all_r)
```
32 Doppelgangers within Ross et al.

```{r}
table(doppel_all_r$PPCC_df$DoppelgangerLabel)
```

### b) Within Yeoh et al.

```{r}
doppel_y = getPPCCDoppelgangers(all_y, 
                                all_y_meta,
                                do_min_max = TRUE)
```

```{r}
visualisePPCCDoppelgangers(doppel_y)
```

There are 121 Doppelgangers in Yeoh et al.

```{r}
table(doppel_y$PPCC_df$DoppelgangerLabel)
```

### c) Between Ross and Yeoh

```{r}
shared_variables_all =  intersect(
                          rownames(all_r),
                          rownames(all_y))
all = data.frame(cbind(all_r[shared_variables_all,],
                       all_y[shared_variables_all,]))
all_meta = data.frame(rbind(all_r_meta, all_y_meta))
doppel_ry = getPPCCDoppelgangers(all, 
                                 all_meta,
                                 do_min_max = TRUE)
```
Only 41 Doppelgangers between AllenData and MaryData.

```{r}
table(doppel_ry$PPCC_df$DoppelgangerLabel)
```

```{r}
doppel_ry$PPCC_df[doppel_ry$PPCC_df$DoppelgangerLabel=="Doppelganger",]
```
```{r}
doppel_ry$cut_off
```

```{r}
visualisePPCCDoppelgangers(doppel_ry)
```
## 3) Batch Imbalance on Doppelganger Identification

Since sva:Combat is used in the doppelgangerIdentifier and batch imbalance affects the efficacy of batch correction, we will now be exploring the impacts of batch imbalance on doppelgangerIdentifier's performance. 

Ross and Yeoh do not have the same sample size. To test the impact of batch imbalance, we test the case with imbalance (the previous result) and the case with no imbalance (oversampling of Ross).

```{r}
table(all_meta$Batch)
```

### a) If we do not balance the batches before doppelganger identification

We identify 41 doppelgangers can seen in the previous section.

### b) If we balance the batches before we identify doppelgangers

We will now over sample Ross

```{r}
oversample = oversample_batch(all, all_meta)
all_over = oversample$raw_data
all_meta_over = oversample$meta_data
table(all_meta_over$Class, all_meta_over$Batch)
```

```{r}
doppel_ry_over = getPPCCDoppelgangers(all_over, 
                                      all_meta_over,
                                      do_min_max = TRUE)
```

Remove extra pairs
```{r}
doppel_ry_over = remove_all_dup(doppel_ry_over)
```

When batches are balanced, 22 doppelgangers were identified.

```{r}
table(doppel_ry_over$PPCC_df$DoppelgangerLabel)
```

No additional doppelgangers were identified after batch balancing

```{r}
dfSetDifference(
  doppel_ry_over$PPCC_df
  [doppel_ry_over$PPCC_df$DoppelgangerLabel=="Doppelganger",],
  doppel_ry$PPCC_df
  [doppel_ry$PPCC_df$DoppelgangerLabel=="Doppelganger",]
  )
```

19 more doppelgangers were identified in the batch imbalance case

```{r}
dfSetDifference(
  doppel_ry$PPCC_df
  [doppel_ry$PPCC_df$DoppelgangerLabel=="Doppelganger",],
  doppel_ry_over$PPCC_df
  [doppel_ry_over$PPCC_df$DoppelgangerLabel=="Doppelganger",]
)
```

```{r}
visualisePPCCDoppelgangers(doppel_ry_over)
```


```{r}
summary(doppel_ry$PPCC_df$PPCC)
```

```{r}
summary(doppel_ry_over$PPCC_df$PPCC)
```

When batches were balanced before doppelganger identification, PPCC decreased slightly. 

### c) Batch imbalance on batch correction efficacy

```{r}
if (!"ggfortify" %in% installed.packages()){
  install.packages("ggfortify")
}
library(ggfortify)
```

#### Before Batch Correction
```{r}
pca_res = prcomp(t(all), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

#### With batch imbalance

```{r}
pca_res = prcomp(t(doppel_ry$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

#### Without batch imbalance

```{r}
pca_res = prcomp(t(doppel_ry_over$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = all_meta, colour = 'Class', shape="Batch")
```

Output information from above into excel for planning purposes.

```{r}
output_cross_batch_planning_xlsx(
  meta_data_df = all_meta,
  doppel_results_bal = doppel_ry_over,
  doppel_results_unbal = doppel_ry,
  filename = "all_planning.xlsx"
)
```

## 4) Testing the doppelganger effect of identified doppelgangers

Here we examine the inflationary effects of the doppelganger identified in the balanced case

```{r}
veri_results_all = verifyDoppelgangers(
  file.path(experiment_plans_dir, "all_experiment_plan.csv"),
  doppel_ry_over$Processed_data,
  all_meta,
  do_batch_corr = FALSE,
  size_of_val_set = 10,
  k=7
)
```

```{r fig.width=10, fig.height=5}
originalTrainValidNames =  c("Doppel_0", "Doppel_2", "Doppel_4", "Doppel_5","Pos_Con_5", "Neg_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel","4 Doppel", "5 Doppel","Pos Con 5", "Neg Con")

visualiseVerificationResults(veri_results_all,
                originalTrainValidNames,
                newTrainValidNames)
```

```{r}
check_veri_stats(
  meta_data_df = all_meta,
  veri_results = veri_results_all,
  doppel_results = doppel_ry_over
)
```

Here we examine the inflationary effects of additional doppelgangers identified in the unbalanced case.

```{r}
veri_results_all_add = verifyDoppelgangers(
  file.path(experiment_plans_dir, "all_experiment_plan_additional.csv"),
  doppel_ry_over$Processed_data,
  all_meta,
  do_batch_corr = FALSE,
  size_of_val_set = 10,
  k=7
)
```

```{r fig.width=10, fig.height=5}
originalTrainValidNames_add =  c("Doppel_0", "E2A_2_R", "E2A_13_R", "Pos_Con_1", "Neg_Con")

newTrainValidNames_add =  c("0 Doppel", "E2A_2_R\n(Unbal)","E2A_13_R\n(Unbal)", "Pos Con 1", "Neg Con")

visualiseVerificationResults(veri_results_all_add,
                originalTrainValidNames_add,
                newTrainValidNames_add)
```

```{r}
check_veri_stats(
  meta_data_df = all_meta,
  veri_results = veri_results_all_add,
  doppel_results = doppel_ry
)
```
```{r}
check_veri_stats(
  meta_data_df = all_meta,
  veri_results = veri_results_all_add,
  doppel_results = doppel_ry_over
)
```