---
title: "Exploring Doppelgangers in the CCLE Dataset"
output: rmarkdown::github_document
author: "Wang Li Rong"
---

# 0. Importing Libraries and Useful Functions

## a) Importing libraries

```{r}
if (!("doppelgangerIdentifier" %in% installed.packages())){
  install.packages('devtools')
  library(devtools)
  install_github('lr98769/doppelgangerIdentifier')
}
library("doppelgangerIdentifier")
if (!("openxlsx") %in% installed.packages()){
  install.packages('openxlsx')
}
library(openxlsx)
```

## b) Importing Functions

```{r}
functions_dir = "../functions"
source(file.path(functions_dir, "process_dataframes.R"))
source(file.path(functions_dir, "process_vectors.R"))
source(file.path(functions_dir, "process_lists.R"))
source(file.path(functions_dir, "output_functions.R"))
source(file.path(functions_dir, "planning_functions.R"))
```

## c) File Paths

```{r}
dataset_url = "https://depmap.org/portal/download/api/download?file_name=ccle%2Fccle_2019%2FCCLE_RNAseq_rsem_genes_tpm_20180929.txt.gz&bucket=depmap-external-downloads"

dataset_dir = "../data"
processed_dataset_dir = "../cleaned_data"
planning_data_dir = "../planning_data"
experiment_plans_dir = "../experiment_plans"
images_dir = "../images"

dataset_filepath = file.path(dataset_dir, "CCLE_RNAseq_rsem_genes_tpm_20180929.txt.gz")

processed_ccle_filepath = file.path(processed_dataset_dir, 
                                    "ccle.rds")
processed_ccle_meta_filepath = file.path(processed_dataset_dir, 
                                    "ccle_meta.rds")

processed_lymph_lung_filepath = file.path(processed_dataset_dir,
                                          "lymph_lung.rds")
processed_lymph_lung_meta_filepath = file.path(processed_dataset_dir,
                                          "lymph_lung_meta.rds")
planning_data_lymph_lung_filepath = file.path(planning_data_dir,
                                              "lymph_lung_plan.xlsx")
ex_plan_lymph_lung_xlsx_filepath = file.path(experiment_plans_dir,
                                        "lymph_lung_ex_plan.xlsx")
ex_plan_lymph_lung_csv_filepath = file.path(experiment_plans_dir,
                                        "lymph_lung_ex_plan.csv")

processed_large_upper_filepath = file.path(processed_dataset_dir,
                                          "large_upper.rds")
processed_large_upper_meta_filepath = file.path(processed_dataset_dir,
                                          "large_upper_meta.rds")
planning_data_large_upper_filepath = file.path(planning_data_dir,
                                              "large_upper_plan.xlsx")
ex_plan_large_upper_xlsx_filepath = file.path(experiment_plans_dir,
                                        "large_upper_ex_plan.xlsx")
ex_plan_large_upper_csv_filepath = file.path(experiment_plans_dir,
                                        "large_upper_ex_plan.csv")

ddi_image_filepath= file.path(images_dir, "ccle_ddi.tiff")
dv_image_filepath= file.path(images_dir, "ccle_dv.tiff")

```

# d) Ensure directories and the proper files have been downloaded
```{r}
directories = c(
  dataset_dir, processed_dataset_dir, planning_data_dir, 
  experiment_plans_dir, images_dir
)
for (directory in directories){
  if (!file.exists(directory)){
    dir.create(directory)
  }
}
if (!file.exists(dataset_filepath)){
  download.file(dataset_url, dataset_filepath, mode = "wb")
}

```
# 1. Importing the Dataset

Data Set Statistics:
- 57820 Genes
- 1019 Samples

```{r}
if (!file.exists(processed_ccle_filepath)){
  ccle = read.table(dataset_filepath, row.names = 1)
  ccle = set_first_row_as_header(ccle)
  ccle$transcript_ids = NULL
  ccle[] = lapply(ccle,as.numeric)
  saveRDS(ccle, processed_ccle_filepath)
} else {
  ccle = readRDS(processed_ccle_filepath)
}
```

Getting meta data for CCLE

```{r}
if (!file.exists(processed_ccle_meta_filepath)){
  ccle_meta = generate_meta_data(ccle)
  saveRDS(ccle_meta, processed_ccle_meta_filepath)
} else {
  ccle_meta = readRDS(processed_ccle_meta_filepath)
}
```

Analysing classes in CCLE

```{r}
View(table(ccle_meta$Class))
```

# 2. Finding PPCC Doppelgangers Within CCLE

We chose 2 sets of cancer pairs to look for doppelgangers. 

## a) "Haematopoietic and Lymphoid Tissue" and "Lung" Cancer

This cancer pair was chosen for its large size.

There are 173 HAEMATOPOIETIC_AND_LYMPHOID_TISSUE samples and 188 LUNG samples.

```{r}
if (!file.exists(processed_lymph_lung_meta_filepath)){
  lymph_lung_meta = ccle_meta[ccle_meta$Class %in% c("HAEMATOPOIETIC_AND_LYMPHOID_TISSUE","LUNG"),]
  saveRDS(lymph_lung_meta, processed_lymph_lung_meta_filepath)
} else {
  lymph_lung_meta = readRDS(processed_lymph_lung_meta_filepath)
}

if (!file.exists(processed_lymph_lung_filepath)){
  lymph_lung = ccle[, row.names(lymph_lung_meta)]
  saveRDS(lymph_lung, processed_lymph_lung_filepath)
} else {
  lymph_lung = readRDS(processed_lymph_lung_filepath)
}
table(lymph_lung_meta$Class)
```
Let's observe if there are any doppelgangers within this data.

```{r}
start_time = Sys.time()
lymph_lung_doppel = getPPCCDoppelgangers(
  raw_data = lymph_lung,
  meta_data = lymph_lung_meta,
  do_batch_corr = FALSE,
  do_min_max = TRUE
)
end_time = Sys.time()
end_time - start_time
```

```{r fig.width=10, fig.height=5}
visualisePPCCDoppelgangers(lymph_lung_doppel) + ggtitle(label="PPCC Doppelganger Identification", subtitle="HAEMATOPOIETIC_AND_LYMPHOID_TISSUE + LUNG") + theme(plot.subtitle = element_text(hjust = 0.5))
```

```{r}
# There are 692 Doppelgangers
table(lymph_lung_doppel$PPCC_df$DoppelgangerLabel)
```

```{r}
# There are 181 samples with at least 1 other doppelganger partner
lymph_lung_doppelganger_df = lymph_lung_doppel$PPCC_df[
  lymph_lung_doppel$PPCC_df$DoppelgangerLabel == "Doppelganger",]
length(union(lymph_lung_doppelganger_df$Sample1,
             lymph_lung_doppelganger_df$Sample2))
```

```{r}
lymph_lung_doppelganger_df
```

## b) "Large Intestine" and "Upper Aerodigestive Tract" Cancer

 There are 56 LARGE_INTESTINE samples and 31 UPPER_AERODIGESTIVE_TRACT  samples.

```{r}
if (!file.exists(processed_large_upper_meta_filepath)){
  large_upper_meta = ccle_meta[ccle_meta$Class %in% c("LARGE_INTESTINE","UPPER_AERODIGESTIVE_TRACT"),]
  saveRDS(large_upper_meta, processed_large_upper_meta_filepath)
} else {
  large_upper_meta = readRDS(processed_large_upper_meta_filepath)
}

if (!file.exists(processed_large_upper_filepath)){
  large_upper = ccle[, row.names(large_upper_meta)]
  saveRDS(large_upper, processed_large_upper_filepath)
} else {
  large_upper = readRDS(processed_large_upper_filepath)
}
table(large_upper_meta$Class)
```

Let's observe if there are any doppelgangers within this data.

```{r}
start_time = Sys.time()
large_upper_doppel = getPPCCDoppelgangers(
  raw_data = large_upper,
  meta_data = large_upper_meta,
  do_batch_corr = FALSE,
  do_min_max = TRUE
)
end_time = Sys.time()
end_time - start_time
```
```{r fig.width=10, fig.height=5}
visualisePPCCDoppelgangers(large_upper_doppel) + ggtitle(label="PPCC Doppelganger Identification", subtitle="LARGE_INTESTINE + UPPER_AERODIGESTIVE_TRACT") + theme(plot.subtitle = element_text(hjust = 0.5))
```

```{r}
# 9 of the samples are doppelganger samples with at least one other sample
large_upper_doppelganger_df = large_upper_doppel$PPCC_df[large_upper_doppel$PPCC_df$DoppelgangerLabel == "Doppelganger",]
length(union(large_upper_doppelganger_df$Sample1,
             large_upper_doppelganger_df$Sample2))
```

```{r}
large_upper_doppelganger_df
```

# 3. Verifying the functionality of identified doppelgangers

## a) "Haematopoietic and Lymphoid Tissue" and "Lung" Cancer

Generating csv file with necessary data for experiment planning

```{r}
lymph_lung_planning_list = generate_planning_dataframes(
  meta_data_df = lymph_lung_meta,
  doppel_results = lymph_lung_doppel)
output_list_of_df_to_excel(
  list_to_save=lymph_lung_planning_list,
  filename = planning_data_lymph_lung_filepath
)
```

Generating list of doppelganger validation samples

```{r}
lymph_lung_doppel_sample_in_valid = find_largest_valid_doppel(
  planning_list = lymph_lung_planning_list)

lymph_lung_doppel_sample_in_valid
```

Highlighting doppelganger validation samples in the excel file for a visual check

```{r}
output_highlighted_planning_file(
  planning_list = lymph_lung_planning_list,
  highlighted_filename = planning_data_lymph_lung_filepath,
  doppel_samples_in_valid = lymph_lung_doppel_sample_in_valid
)
```

Generate experiment plan

```{r}
# Select the first 25 and last 25 doppel samples 
valid_doppel = c(head(lymph_lung_doppel_sample_in_valid, 25),
                   tail(lymph_lung_doppel_sample_in_valid, 25))
lymph_lung_experiment_plan = generate_experiment_plan(
  planning_list = lymph_lung_planning_list,
  valid_doppel = valid_doppel,
  num_samples_added = 10)
```
Output the experiment plan to an excel file (For a visual check)

```{r}
highlight_style_list = list(
  train_doppel = createStyle(fgFill = "#d0adff"),
  train_non_doppel = createStyle(fgFill = "#ffffff"),
  valid_doppel = createStyle(fgFill = "#8d29ff"),
  valid_non_doppel = createStyle(fgFill = "#bdbdbd"),
  valid_non_doppel_extra = createStyle(fgFill = "#ffffff")
)

output_experiment_plan(
  experiment_plan = lymph_lung_experiment_plan,
  planning_list = lymph_lung_planning_list,
  ex_plan_csv_fp = ex_plan_lymph_lung_csv_filepath,
  ex_plan_xlsx_fp = ex_plan_lymph_lung_xlsx_filepath,
  highlight_style_list = highlight_style_list)
```
Verifying the functionality of PPCC data doppelgangers

```{r}
lymph_lung_veri = verifyDoppelgangers(
  experiment_plan_filename = ex_plan_lymph_lung_csv_filepath,
  raw_data = lymph_lung,
  meta_data = lymph_lung_meta,
  feature_set_portion = 0.1,
  do_batch_corr = FALSE,
  k=17,
  size_of_val_set = 50
)
```

Visualise the results

```{r fig.width=10, fig.height=5}
visualiseVerificationResults(lymph_lung_veri) + ggtitle(label="Accuracy of KNN Models", subtitle="HAEMATOPOIETIC_AND_LYMPHOID_TISSUE + LUNG\nValidation Set Size = 50 Samples") + theme(plot.subtitle = element_text(hjust = 0.5))
```

## b) "Large Intestine" and "Upper Aerodigestive Tract" Cancer

Generating csv file with necessary data for experiment planning

```{r}
large_upper_planning_list = generate_planning_dataframes(
  meta_data_df = large_upper_meta,
  doppel_results = large_upper_doppel)
```

Get largest possible set of doppelganger samples in validation

```{r}
large_upper_valid_doppel = find_largest_valid_doppel(
  planning_list = large_upper_planning_list
)
large_upper_valid_doppel
```

Get experiment plan with this list of doppelganger samples in validation

```{r}
large_upper_experiment_plan = generate_experiment_plan(
  planning_list = large_upper_planning_list,
  valid_doppel = large_upper_valid_doppel,
  num_samples_added = 1
)
```

Save experiment plans

```{r}
output_experiment_plan(
  experiment_plan = large_upper_experiment_plan,
  planning_list = large_upper_planning_list,
  ex_plan_csv_fp = ex_plan_large_upper_csv_filepath,
  ex_plan_xlsx_fp = ex_plan_large_upper_xlsx_filepath,
  highlight_style_list = highlight_style_list
)
```
Verify functionality of PPCC data doppelgangers

```{r}
large_upper_veri = verifyDoppelgangers(
  experiment_plan_filename = ex_plan_large_upper_csv_filepath,
  raw_data = large_upper,
  meta_data = large_upper_meta,
  feature_set_portion = 0.1,
  do_batch_corr = FALSE,
  k=9,
  size_of_val_set = 10
)
```

View verification results

```{r fig.width=10, fig.height=5}
visualiseVerificationResults(large_upper_veri) + ggtitle(label="Accuracy of KNN Models", subtitle="LARGE_INTESTINE + UPPER_AERODIGESTIVE_TRACT\nValidation Set Size = 10 Samples") + theme(plot.subtitle = element_text(hjust = 0.5))
```

# 4. Arrange Plots

## a) DD Identification

```{r fig.height=6, fig.width=8}
if (!"ggpubr" %in% installed.packages()){
  install.packages("ggpubr")
}
library(ggpubr)

# Arrange both plots
ccle_ddi_plots = ggarrange(
          ggpar(visualisePPCCDoppelgangers(lymph_lung_doppel) +
            coord_cartesian(ylim = c(0, 1)),
            title="lymph_lung"),
          ggpar(visualisePPCCDoppelgangers(large_upper_doppel) +
            coord_cartesian(ylim = c(0, 1)),
            title="large_upper")+
          rremove("ylab") +
          rremove("y.ticks")+
          rremove("y.text"),
          align='h',
          labels=c('A', 'B'),
          common.legend = T,
          legend = "bottom"
          )
# Add big title above
ccle_ddi_plots = annotate_figure(
  ccle_ddi_plots,
  top = text_grob("PPCC DD Identification",
  color = "black",
  face = "bold",
  size = 18))
ccle_ddi_plots
```

```{r}
# Output
ggsave(filename = ddi_image_filepath, width = 8, height = 6, device='tiff')
```

## b) DD Verification

```{r fig.width=8, fig.height=10}
if (!"ggpubr" %in% installed.packages()){
  install.packages("ggpubr")
}
library(ggpubr)

# Arrange both plots
ccle_dv_plots = ggarrange(
  ggpar(visualiseVerificationResults(lymph_lung_veri,
    c("Doppel_0","Doppel_10", "Doppel_20", "Doppel_30", "Doppel_40","Doppel_50", "Pos_Con_50", "Neg_Con"),
    c("0 Doppel", "10 Doppel", "20 Doppel", "30 Doppel", "40 Doppel", "50 Doppel", "50 Pos Con", "Neg Con")) +
    coord_cartesian(ylim = c(0.3, 1.05)),
    title="lymph_lung") +
    scale_y_continuous(breaks=seq(0.3,1,0.1)) +
    labs(subtitle="311 Training, 50 Validation") + 
      theme(plot.subtitle=element_text(size=11, hjust=0.5, color="#2e2e2e")) + 
      theme(plot.title=element_text(size=12, hjust=0.5, color="black", face="bold")),
  ggpar(visualiseVerificationResults(large_upper_veri,
    c("Doppel_0","Doppel_1", "Doppel_2", "Doppel_3", "Doppel_4", "Doppel_5", "Pos_Con_5", "Neg_Con"),
    c("0 Doppel", "1 Doppel", "2 Doppel", "3 Doppel", "4 Doppel", "5 Doppel", "5 Pos Con", "Neg Con")
    ) +
    coord_cartesian(ylim = c(0.3, 1.05)),
    title="large_upper") +
    scale_y_continuous(breaks=seq(0.3,1,0.1)) +
    labs(subtitle="77 Training, 10 Validation") + 
      theme(plot.subtitle=element_text(size=11, hjust=0.5, color="#2e2e2e")) + 
      theme(plot.title=element_text(size=12, hjust=0.5, color="black", face="bold")),
  align='v',
  labels=c('A', 'B'),
  common.legend = T,
  legend = "bottom",
  nrow=2
)
# Add big title above
ccle_dv_plots = annotate_figure(
  ccle_dv_plots,
  top = text_grob("Functional Doppelganger Verification",
  color = "black",
  face = "bold",
  size = 18))
ccle_dv_plots

```

```{r}
# Output
ggsave(filename = dv_image_filepath, width = 8, height = 10, device='tiff')
```

