---
title: "The Doppelganger Effect in Gene Expression Data sets"
output: rmarkdown::github_document
author: "Wang Li Rong"
---

# 1. Exploring the (Duchenne Muscular Dystrophy) DMD Datasets

## 0) Import packages

```{r Import & Install Packages, include=FALSE}
if (!("doppelgangerIdentifier" %in% installed.packages())){
  install.packages('devtools')
  library(devtools)
  install_github('lr98769/doppelgangerIdentifier')
}
library("doppelgangerIdentifier")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!("biomaRt" %in% installed.packages())){
  BiocManager::install("biomaRt")
}
library("biomaRt")
```

File paths
```{r}
data_dir = "../data"
cleaned_data_dir = "../cleaned_data"
images_dir = "../images"
planning_data_dir = "../planning_data"
experiment_plans_dir = "../experiment_plans"
```

Load functions for pre-processing and analysis:
```{r}
functions_dir = "../functions"
source(file.path(functions_dir,"preprocessing_functions.R"))
source(file.path(functions_dir,"analysis_functions.R"))
source(file.path(functions_dir,"planning_functions.R"))
source(file.path(functions_dir,"process_vectors.R"))
source(file.path(functions_dir,"process_lists.R"))
source(file.path(functions_dir,"visualisation_code.R"))
```

## 1) Import the DMD datasets

HaslettData
- 12 DMD Samples
- 12 Normal Samples

```{r}
# Converts the affy probes to ensemble id
dmd_h = getDataFile(
  filename= file.path(data_dir, "DMD-HaslettData.csv"),
  affy_attribute = "affy_hg_u95av2",
  batch_name = "H")
#Forming The metadata dataframe 
dmd_h_meta = getMetaDataDataframe(df = dmd_h, batch_name = "H")
```

PescatoriData
- 22 DMD Samples
- 14 Normal Samples

```{r}
dmd_p = getDataFile(filename = 
  file.path(data_dir, "DMD-PescatoriData.csv"),
  affy_attribute = "affy_hg_u133a",
  batch_name = "P")
#Forming The metadata dataframe 
dmd_p_meta = getMetaDataDataframe(df = dmd_p, batch_name = "P")
```

## 2) Identify doppelgangers within and between both batches

### a) Doppelgangers within HaslettData

33 Doppelgangers found within HaslettData.
```{r}
doppel_h = getPPCCDoppelgangers(dmd_h, dmd_h_meta, do_min_max = TRUE)
table(doppel_h$PPCC_df$DoppelgangerLabel)
```

```{r}
visualisePPCCDoppelgangers(doppel_h)
```

### b) Doppelgangers within PescatoriData

There are 8 doppelgangers within PescatoriData.

```{r}
doppel_p = getPPCCDoppelgangers(dmd_p, dmd_p_meta, do_min_max = TRUE)
table(doppel_p$PPCC_df$DoppelgangerLabel)
```

```{r}
visualisePPCCDoppelgangers(doppel_p)
```

### b) Doppelgangers Between Both Batches

Combining both data sets into one 

```{r}
dmd_meta = data.frame(rbind(dmd_h_meta, dmd_p_meta))
#8813 variables in common
shared_variables_dmd = intersect(rownames(dmd_h), rownames(dmd_p)) 
dmd = data.frame(cbind(dmd_h[shared_variables_dmd, ], 
                       dmd_p[shared_variables_dmd,]))
```

Finding doppelgangers between both datasets

```{r}
doppel_hp = getPPCCDoppelgangers(dmd, dmd_meta, do_min_max = TRUE)
```
54 Doppelgangers between HaslettData and PescatoriData.

```{r}
table(doppel_hp$PPCC_df$DoppelgangerLabel)
```

```{r}
visualisePPCCDoppelgangers(doppel_hp)
```

It seems that NOR_12_H is negatively correlated with all samples of batch P. This seems to be the result of NOR_12_H having abnormally low values and batch correction converting these low values to negative values. 

```{r}
doppel_hp$PPCC_df[doppel_hp$PPCC_df$Sample1=="NOR_12_H" &
                    doppel_hp$PPCC_df$PPCC<0,]
```
```{r fig.width=8, fig.height=5}
doppel_hp_rem = remove_sample_from_doppel_result(
  doppel_result = doppel_hp,
  sample = "NOR_12_H"
)
visualisePPCCDoppelgangers(doppel_hp_rem)
```

## 3) Batch Effect on Doppelganger Identification

### a) With batch imbalances
The above results show the case where batch imbalances exist.

```{r}
table(dmd_meta$Batch)
```

### b) Without batch imbalance

Here we will be oversampling the data set before doppelganger identification resolve batch imbalances. 

```{r}
# Oversampling data
oversample = oversample_batch(dmd, dmd_meta)
dmd_over = oversample$raw_data
dmd_meta_over = oversample$meta_data
table(dmd_meta_over$Batch)
```

```{r}
table( dmd_meta_over$Class, dmd_meta_over$Batch)
```


```{r}
doppel_hp_over = getPPCCDoppelgangers(dmd_over, dmd_meta_over,
                                      do_min_max = TRUE)
# Remove excess sample pairs
doppel_hp_over = remove_all_dup(doppel_hp_over)
```

When there is an absence of batch imbalance, lesser doppelgangers were identified

```{r}
table(doppel_hp_over$PPCC_df$DoppelgangerLabel)
```

2 more pairs were identified

```{r}
dfSetDifference(
  doppel_hp_over$PPCC_df[doppel_hp_over$PPCC_df$DoppelgangerLabel=="Doppelganger", ],
  doppel_hp$PPCC_df[doppel_hp$PPCC_df$DoppelgangerLabel=="Doppelganger", ]
)
```

9 pairs were not identified the the batch balanced case

```{r}
dfSetDifference(
  doppel_hp$PPCC_df[doppel_hp$PPCC_df$DoppelgangerLabel=="Doppelganger", ],
  doppel_hp_over$PPCC_df[doppel_hp_over$PPCC_df$DoppelgangerLabel=="Doppelganger", ]
 )
```

The presence of batch imbalances decreased PPCC slightly in this case. 

```{r}
doppel_hp_over_rem = remove_sample_from_doppel_result(
  doppel_result = doppel_hp_over,
  sample = "NOR_12_H"
)
visualisePPCCDoppelgangers(doppel_hp_over_rem)
```

```{r}
summary(doppel_hp$PPCC_df$PPCC)
```

```{r}
summary(doppel_hp_over$PPCC_df$PPCC)
```

With no batch imbalances, PPCC increased slightly.

```{r}
print(paste("Unbalanced cut-off:", doppel_hp_rem$cut_off))
print(paste("Balanced cut-off:", doppel_hp_over_rem$cut_off))
```

### c) Batch imbalance on batch correction efficacy

```{r}
if (!"ggfortify" %in% installed.packages()){
  install.packages("ggfortify")
}
library(ggfortify)
```

#### Before Batch Correction
```{r}
pca_res = prcomp(t(dmd), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```
```{r}
autoplot(x=2, y=3, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```

#### With batch imbalance

```{r}
pca_res = prcomp(t(doppel_hp$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```

#### Without batch imbalance

```{r}
pca_res = prcomp(t(doppel_hp_over$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```
```{r}
autoplot(x=2, y=3, pca_res, data = dmd_meta, colour = 'Class', shape="Batch")
```

Output the doppelgangers for easier planning

```{r}
output_cross_batch_planning_xlsx(
  meta_data_df = dmd_meta,
  doppel_results_bal = doppel_hp_over_rem,
  doppel_results_unbal = doppel_hp_rem,
  filename = "dmd_planning.xlsx"
)
```

## 4) Testing the Doppelganger Effect
Here we test the doppelganger effects of PPCC data doppelgangers identified in the balanced case

```{r}
veri_results_dmd =  verifyDoppelgangers(
  file.path(experiment_plans_dir,"dmd_experiment_plan.csv"),
  doppel_hp_over$Processed_data,
  dmd_meta,
  do_batch_corr = FALSE,
  size_of_val_set = 10,
  k=5,
  feature_set_portion = 0.01,
  num_random_feature_sets = 20
)
```

```{r fig.height=5, fig.width=10}
originalTrainValidNames =  c("Doppel_0", "Doppel_2","Doppel_4","Doppel_6", "Doppel_8", "Doppel_10", "Pos_Con_10", "Neg_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel", "4 Doppel", "6 Doppel","8 Doppel", "10 Doppel", "10 Pos Con", "Neg Con")

visualiseVerificationResults(veri_results_dmd,
                originalTrainValidNames,
                newTrainValidNames)
```

```{r fig.height=5, fig.width=10}
originalTrainValidNames =  c("Doppel_0", "Doppel_2","Doppel_4","Doppel_6", "Doppel_8", "Doppel_10", "Pos_Con_10", "Neg_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel", "4 Doppel", "6 Doppel","8 Doppel", "10 Doppel", "10 Pos Con", "Neg Con")

visualiseVerificationResults2(veri_results_dmd,
                originalTrainValidNames,
                newTrainValidNames)
```

```{r}
check_veri_stats(
  meta_data_df = dmd_meta,
  veri_results = veri_results_dmd,
  doppel_results = doppel_hp_over_rem
)
```

Here we test the doppelganger effects of the additionally identified PPCC data doppelgangers 

```{r}
veri_results_dmd_add =  verifyDoppelgangers(
  file.path(experiment_plans_dir,"dmd_experiment_plan_additional.csv"),
  doppel_hp_over$Processed_data,
  dmd_meta,
  do_batch_corr = FALSE,
  size_of_val_set = 10,
  k=5,
  feature_set_portion = 0.01,
  num_random_feature_sets = 20
)
```

```{r fig.height=5, fig.width=10}
originalTrainValidNames_add =  c("Doppel_0", "DMD_13_P","NOR_2_P","NOR_7_P", "Pos_Con_3", "Neg_Con")

newTrainValidNames_add =  c("0 Doppel", "DMD_13_P\n(Unbal)","NOR_2_P\n(Unbal)","NOR_7_P\n(Unbal)", "3 Pos Con", "Neg Con")

visualiseVerificationResults(veri_results_dmd_add,
                originalTrainValidNames_add,
                newTrainValidNames_add)
```

```{r fig.height=5, fig.width=10}
originalTrainValidNames_add =  c("Doppel_0", "DMD_13_P","NOR_2_P","NOR_7_P", "Pos_Con_3", "Neg_Con")

newTrainValidNames_add =  c("0 Doppel", "DMD_13_P\n(Unbal)","NOR_2_P\n(Unbal)","NOR_7_P\n(Unbal)", "3 Pos Con", "Neg Con")

visualiseVerificationResults2(veri_results_dmd_add,
                originalTrainValidNames_add,
                newTrainValidNames_add)
```

```{r}
check_veri_stats(
  meta_data_df = dmd_meta,
  veri_results = veri_results_dmd_add,
  doppel_results = doppel_hp
)
```


```{r}
check_veri_stats(
  meta_data_df = dmd_meta,
  veri_results = veri_results_dmd_add,
  doppel_results = doppel_hp_over
)
```

