---
title: "The Doppelganger Effect in Biomedical Data sets"
output: rmarkdown::github_document
author: "Wang Li Rong"
---

# 2. Exploring the Leukemia Datasets

## 0) Import packages

```{r Import & Install Packages, include=FALSE}
if (!("doppelgangerIdentifier" %in% installed.packages())){
  install.packages('devtools')
  library(devtools)
  install_github('lr98769/doppelgangerIdentifier')
}
library("doppelgangerIdentifier")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!("biomaRt" %in% installed.packages())){
  BiocManager::install("biomaRt")
}
library("biomaRt")
```

File paths
```{r}
data_dir = "../data"
cleaned_data_dir = "../cleaned_data"
images_dir = "../images"
planning_data_dir = "../planning_data"
experiment_plans_dir = "../experiment_plans"
```

Load functions for pre-processing and analysis:
```{r}
functions_dir = "../functions"
source(file.path(functions_dir,"preprocessing_functions.R"))
source(file.path(functions_dir,"analysis_functions.R"))
source(file.path(functions_dir,"planning_functions.R"))
source(file.path(functions_dir,"process_vectors.R"))
source(file.path(functions_dir,"process_lists.R"))
source(file.path(functions_dir,"visualisation_code.R"))
```


## 1) Import the datasets
ArmstrongData
- 24 ALL
- 24 AML

```{r}
leuk_A = getDataFile(
  filename = file.path(data_dir, "Leuk-ArmstrongData.csv"),
  affy_attribute = "affy_hg_u95av2",
  batch_name = "A")
leuk_A_meta = getMetaDataDataframe(df = leuk_A, batch_name = "A")
```

GolubData
- 47 ALL
- 25 AML
```{r}
leuk_G = getGolubDataFile()
leuk_G_meta = getMetaDataDataframe(df = leuk_G,
                                   batch_name = "G")

```

## 2) Identifying PPCC Data Doppelgangers

### a) Within the Armstrong Data

```{r}
doppel_a = getPPCCDoppelgangers(leuk_A, 
                                leuk_A_meta,
                                do_min_max = TRUE)
```

```{r}
visualisePPCCDoppelgangers(doppel_a)
```

```{r}
doppel_a$cut_off
```

There are 38 Doppelgangers. 

```{r}
table(doppel_a$PPCC_df$DoppelgangerLabel) 
```

### b) Within the Golub Data

```{r}
doppel_g = getPPCCDoppelgangers(leuk_G, 
                                leuk_G_meta,
                                do_min_max = TRUE)
```

```{r}
visualisePPCCDoppelgangers(doppel_g)
```

```{r}
doppel_g$cut_off
```

There are 8 doppelgangers in total. 

```{r}
table(doppel_g$PPCC_df$DoppelgangerLabel)
```

### c) Between Armstrong and Golub

```{r}
shared_variables_leuk = intersect(
  rownames(leuk_A), rownames(leuk_G)
)
leuk = data.frame(cbind(leuk_A[shared_variables_leuk,],
                        leuk_G[shared_variables_leuk,]))
leuk_meta = data.frame(rbind(leuk_A_meta, leuk_G_meta))
doppel_ag = getPPCCDoppelgangers(leuk, 
                                 leuk_meta,
                                 do_min_max = TRUE)
```

There are 6 doppelgangers between ArmstringData and GolubData.

```{r}
table(doppel_ag$PPCC_df$DoppelgangerLabel)
```

```{r}
doppel_ag$PPCC_df[doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",]
```

```{r}
visualisePPCCDoppelgangers(doppel_ag)
```
## 3) Batch effect on doppelganger identification

### a) If we do not balance the batches prior to doppelganger identification

We get the above result. 6 Doppelgangers will be identified.

```{r}
table(leuk_meta$Batch)
```

### b) If we balance the batches before doppelganger identification.

To balance the two batches, we oversample ArmstrongData

```{r}
oversample = oversample_batch(raw_data = leuk, meta_data = leuk_meta)
leuk_over = oversample$raw_data
leuk_meta_over = oversample$meta_data
table(leuk_meta_over$Class, leuk_meta_over$Batch)
```

```{r}
doppel_ag_over = getPPCCDoppelgangers(leuk_over, 
                                      leuk_meta_over,
                                      do_min_max = TRUE)
```
```{r}
doppel_ag_over = remove_all_dup(doppel_ag_over)
```

When we balance the batches before doppelganger identification. 9 doppelgangers were identified (More than previous).

```{r}
table(doppel_ag_over$PPCC_df$DoppelgangerLabel)
```

These 3 pairs were identified as doppelgangers when samples were balanced. 

```{r}
dfSetDifference(
  doppel_ag_over$PPCC_df[
    doppel_ag_over$PPCC_df$DoppelgangerLabel=="Doppelganger",], doppel_ag$PPCC_df[
    doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",])
```

No extra pairs were identified as doppelgangers in the unbalanced case

```{r}
dfSetDifference(
  doppel_ag$PPCC_df[
    doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",],
  doppel_ag_over$PPCC_df[
    doppel_ag_over$PPCC_df$DoppelgangerLabel=="Doppelganger",]
)
```

```{r}
visualisePPCCDoppelgangers(doppel_ag_over)
```

```{r}
summary(doppel_ag$PPCC_df$PPCC)
```

```{r}
summary(doppel_ag_over$PPCC_df$PPCC)
```

Only slight differences (decrease) in PPCC distribution after oversampling.

```{r}
print(paste("Unbalanced cut-off:", doppel_ag$cut_off))
print(paste("Balanced cut-off:", doppel_ag_over$cut_off))
```

### c) Batch imbalance on batch correction efficacy

```{r}
if (!"ggfortify" %in% installed.packages()){
  install.packages("ggfortify")
}
library(ggfortify)
```

#### Before Batch Correction
```{r}
pca_res = prcomp(t(leuk), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

#### With batch imbalance

```{r}
pca_res = prcomp(t(doppel_ag$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```
```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

#### Without batch imbalance

```{r}
pca_res = prcomp(t(doppel_ag_over$Processed_data), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

Output the doppelgangers for easier planning

```{r}
output_cross_batch_planning_xlsx(
  meta_data_df = leuk_meta,
  doppel_results_bal = doppel_ag_over,
  doppel_results_unbal = doppel_ag,
  filename = "leuk_planning.xlsx"
)
```

## 4) Testing for doppelganger effect

Here we test the doppelganger effects of PPCC doppelgangers identified in the balanced case

```{r}
veri_results_leuk = verifyDoppelgangers(
      file.path(experiment_plans_dir, "leuk_experiment_plan.csv"),
      doppel_ag_over$Processed_data,
      leuk_meta,
      do_batch_corr = FALSE,
      k=7,
      size_of_val_set = 10,
      feature_set_portion = 0.01,
      num_random_feature_sets = 20
  )
```
```{r fig.width=10, fig.height=5}
originalTrainValidNames =  c("Doppel_0", "Doppel_2", "Doppel_4", "Doppel_6","Pos_Con_6", "Neg_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel", "4 Doppel","6 Doppel","6 Pos Con", "Neg Con")

visualiseVerificationResults2(veri_results_leuk,
                originalTrainValidNames,
                newTrainValidNames)
```

```{r}
check_veri_stats(
  meta_data_df = leuk_meta,
  veri_results = veri_results_leuk,
  doppel_results = doppel_ag_over
)
```

Here we test the doppelganger effects of PPCC doppelgangers identified in the balanced case but not in the unbalanced cse

```{r}
veri_results_leuk_add = verifyDoppelgangers(
    file.path(experiment_plans_dir, "leuk_experiment_plan_additional.csv"),
    doppel_ag_over$Processed_data,
    leuk_meta,
    do_batch_corr = FALSE,
    k=7,
    size_of_val_set = 10,
    feature_set_portion = 0.01,
    num_random_feature_sets = 20,
)
```

```{r fig.width=10, fig.height=5}
originalTrainValidNames_add =  c("Doppel_0", "ALL_22_G", "ALL_19_G", "Pos_Con_2", "Neg_Con")

newTrainValidNames_add =  c("0 Doppel", "ALL_22_G\n(Bal)", "ALL_19_G\n(Bal)","2 Pos Con", "Neg Con")

visualiseVerificationResults(veri_results_leuk_add,
                originalTrainValidNames_add,
                newTrainValidNames_add)
```

```{r fig.width=10, fig.height=5}
originalTrainValidNames_add =  c("Doppel_0", "ALL_22_G", "ALL_19_G", "Pos_Con_2", "Neg_Con")

newTrainValidNames_add =  c("0 Doppel", "ALL_22_G\n(Bal)", "ALL_19_G\n(Bal)","2 Pos Con", "Neg Con")

visualiseVerificationResults2(veri_results_leuk_add,
                originalTrainValidNames_add,
                newTrainValidNames_add)
```

```{r}
check_veri_stats(
  meta_data_df = leuk_meta,
  veri_results = veri_results_leuk_add,
  doppel_results = doppel_ag_over
)
```

```{r}
check_veri_stats(
  meta_data_df = leuk_meta,
  veri_results = veri_results_leuk_add,
  doppel_results = doppel_ag
)
```
