---
title: "The Doppelganger Effect in Biomedical Data sets"
output:
  html_document:
    df_print: paged
---

# 2. Exploring the Leukemia Datasets

## 0) Import packages

```{r Import & Install Packages, include=FALSE}
if (!("doppelgangerIdentifier" %in% installed.packages())){
  install.packages('devtools')
  library(devtools)
  install_github('lr98769/doppelgangerIdentifier')
}
library("doppelgangerIdentifier")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!("biomaRt" %in% installed.packages())){
  BiocManager::install("biomaRt")
}
library("biomaRt")
```

File paths
```{r}
data_dir = "../data"
cleaned_data_dir = "../cleaned_data"
functions_dir = "../functions"
images_dir = "../images"
output_dir = "../output"
```

Load functions for pre-processing and analysis:
```{r}
source(file.path(functions_dir,"preprocessing_functions.R"))
source(file.path(functions_dir,"analysis_functions.R"))
```


## 1) Import the datasets
ArmstrongData
- 24 ALL
- 24 AML

```{r}
leuk_A = getDataFile(
  filename = file.path(data_dir, "Leuk-ArmstrongData.csv"),
  affy_attribute = "affy_hg_u95av2",
  batch_name = "A")
leuk_A_meta = getMetaDataDataframe(df = leuk_A, batch_name = "A")
```

GolubData
- 47 ALL
- 25 AML
```{r}
leuk_G = getGolubDataFile()
leuk_G_meta = getMetaDataDataframe(df = leuk_G,
                                   batch_name = "G")

```

## 2) Identifying PPCC Data Doppelgangers

### a) Within the Armstrong Data

```{r}
doppel_a = getPPCCDoppelgangers(leuk_A, leuk_A_meta)
```

```{r}
visualisePPCCDoppelgangers(doppel_a)
```

```{r}
doppel_a$cut_off
```

There are 100 Doppelgangers. 

```{r}
table(doppel_a$PPCC_df$DoppelgangerLabel) 
```

### b) Within the Golub Data

```{r}
doppel_g = getPPCCDoppelgangers(leuk_G, leuk_G_meta)
```

```{r}
visualisePPCCDoppelgangers(doppel_g)
```

```{r}
doppel_g$cut_off
```

There are 76 doppelgangers in total. 

```{r}
table(doppel_g$PPCC_df$DoppelgangerLabel)
```

### c) Between Armstrong and Golub

```{r}
shared_variables_leuk = intersect(
  rownames(leuk_A), rownames(leuk_G)
)
leuk = data.frame(cbind(leuk_A[shared_variables_leuk,],
                        leuk_G[shared_variables_leuk,]))
leuk_meta = data.frame(rbind(leuk_A_meta, leuk_G_meta))
doppel_ag = getPPCCDoppelgangers(leuk, leuk_meta)
```

There are 28 doppelgangers between ArmstringData and GolubData.

```{r}
table(doppel_ag$PPCC_df$DoppelgangerLabel)
```

```{r}
doppel_ag$PPCC_df[doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",]
```

```{r}
visualisePPCCDoppelgangers(doppel_ag)
```
## 3) Batch effect on doppelganger identification

### a) If we do not balance the batches prior to doppelganger identification

We get the above result. 28 Doppelgangers will be identified.

```{r}
table(leuk_meta$Batch)
```

### b) If we balance the batches before doppelganger identification.

To balance the two batches, we oversample ArmstrongData

```{r}
oversample = oversample_batch(raw_data = leuk, meta_data = leuk_meta)
leuk_over = oversample$raw_data
leuk_meta_over = oversample$meta_data
table(leuk_meta_over$Class, leuk_meta_over$Batch)
```

```{r}
doppel_ag_over = getPPCCDoppelgangers(leuk_over, leuk_meta_over)
```
```{r}
doppel_ag_over = remove_all_dup(doppel_ag_over)
```

When we balance the batches before doppelganger identification. 35 doppelgangers were identified (More than previous).

```{r}
table(doppel_ag_over$PPCC_df$DoppelgangerLabel)
```

These 7 pairs were identified as doppelgangers when samples were balanced. 

```{r}
dfSetDifference(
  doppel_ag_over$PPCC_df[
    doppel_ag_over$PPCC_df$DoppelgangerLabel=="Doppelganger",], doppel_ag$PPCC_df[
    doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",])
```

```{r}
visualisePPCCDoppelgangers(doppel_ag_over)
```

```{r}
summary(doppel_ag$PPCC_df$PPCC)
```

```{r}
summary(doppel_ag_over$PPCC_df$PPCC)
```

Only slight differences (decrease) in PPCC distribution after oversampling.

### c) Batch imbalance on batch correction efficacy

```{r}
if (!"ggfortify" %in% installed.packages()){
  install.packages("ggfortify")
}
library(ggfortify)
```

#### Before Batch Correction
```{r}
pca_res = prcomp(t(leuk), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

#### With batch imbalance

```{r}
pca_res = prcomp(t(doppel_ag$Batch_corrected), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```
```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

#### Without batch imbalance

```{r}
pca_res = prcomp(t(doppel_ag_over$Batch_corrected), scale. = TRUE)
autoplot(x=1, y=2, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

```{r}
autoplot(x=2, y=3, pca_res, data = leuk_meta, colour = 'Class', shape="Batch")
```

Output information from above into excel for planning purposes.

```{r}
if (!"openxlsx" %in% installed.packages()){
  install.packages("openxlsx")
}
library(openxlsx)
wb = createWorkbook()
addWorksheet(wb, "MetaData")
addWorksheet(wb, "DoppelgangerPairs")
addWorksheet(wb, "ExtraPairs_Bal-Unbal")
addWorksheet(wb, "DoppelgangerSamplesA")
addWorksheet(wb, "DoppelgangerSamplesG")

writeData(wb, 1, leuk_meta)
writeData(wb, 2, doppel_ag_over$PPCC_df[
  doppel_ag_over$PPCC_df$DoppelgangerLabel=="Doppelganger",])
writeData(wb, 3, dfSetDifference(
  doppel_ag_over$PPCC_df[
    doppel_ag_over$PPCC_df$DoppelgangerLabel=="Doppelganger",], doppel_ag$PPCC_df[
    doppel_ag$PPCC_df$DoppelgangerLabel=="Doppelganger",]))
writeData(wb, 4, getDoppelnNonDoppelSamples(
  doppel_result = doppel_ag_over,
  metadata = leuk_meta,
  batchname = "A"))
writeData(wb, 5, getDoppelnNonDoppelSamples(
  doppel_result = doppel_ag_over,
  metadata = leuk_meta,
  batchname = "G"))
saveWorkbook(wb, file = "output/leuk_planning.xlsx", overwrite = TRUE)
```

## 4) Testing for doppelganger effect

```{r}
veri_results_leuk = verifyDoppelgangers(
      file.path(cleaned_data_dir, "leuk_experiment_plan.csv"),
      doppel_ag_over$Batch_corrected,
      leuk_meta,
      do_batch_corr = FALSE,
      k=7,
      size_of_val_set = 10
  )
```
```{r fig.width=10, fig.height=5}
originalTrainValidNames =  c("Doppel_0", "Doppel_2", "Doppel_4","Doppel_6","Doppel_8", "Doppel_10", "Neg_Con", "Pos_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel", "4 Doppel","6 Doppel","8 Doppel", "10 Doppel", "Neg Con", "Pos Con")

visualiseVerificationResults(veri_results_leuk,
                originalTrainValidNames,
                newTrainValidNames)
```
## Extra findings

The importance of balanced batches for verification step

```{r fig.width=10, fig.height=5}
veri_results_leuk_extra = verifyDoppelgangers(
     file.path(cleaned_data_dir, "leuk_experiment_plan.csv"),
      leuk,
      leuk_meta,
      k=7,
      size_of_val_set = 10
  )

originalTrainValidNames =  c("Doppel_0", "Doppel_2", "Doppel_4","Doppel_6","Doppel_8", "Doppel_10", "Neg_Con", "Pos_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel", "4 Doppel","6 Doppel","8 Doppel", "10 Doppel", "Neg Con", "Pos Con")

visualiseVerificationResults(veri_results_leuk_extra,
                originalTrainValidNames,
                newTrainValidNames)

```
## Arranging plots

```{r fig.height=4, fig.width=8}
if (!"ggpubr" %in% installed.packages()){
  install.packages("ggpubr")
}
library(ggpubr)

leuk_plot1 = ggarrange(
          ggpar(visualisePPCCDoppelgangers(doppel_ag) +
            coord_cartesian(ylim = c(0.58, 1)),
            title="Unbalanced Batches"), 
          ggpar(visualisePPCCDoppelgangers(doppel_ag_over) +
            coord_cartesian(ylim = c(0.58, 1)),
            title="Balanced Batches")+ 
          rremove("ylab") +
          rremove("y.ticks")+
          rremove("y.text"), 
          align='h', 
          labels=c('A', 'B'),
          common.legend = T,
          legend = "bottom"
          )
leuk_plot1 = annotate_figure(leuk_plot1, 
                            top = text_grob("Leukaemia PPCC Distribution & Doppelganger identification", 
               color = "black",
               face = "bold", 
               size = 18))
leuk_plot1
ggsave(filename = file.path(images_dir, "leuk_PPCC.tiff"), width = 8, height = 4, device='tiff')
```


```{r fig.width=10, fig.height=5}
originalTrainValidNames =  c("Doppel_0", "Doppel_2", "Doppel_4","Doppel_6","Doppel_8", "Doppel_10", "Neg_Con", "Pos_Con")

newTrainValidNames =  c("0 Doppel", "2 Doppel\n (Only In\n Balanced)", "4 Doppel","6 Doppel","8 Doppel", "10 Doppel", "Neg Con", "10 Pos Con")

leuk_plot2 = visualiseVerificationResults(veri_results_leuk,
                 originalTrainValidNames,
                 newTrainValidNames)
leuk_plot2 = ggpar(leuk_plot2, title="Validation Accuracy of KNN Models (Leukaemia)") +
  font("title", size = 18, color = "black", face = "bold") +
  font("xlab", size = 14, color = "black") +
  font("xylab", size = 14, color = "black") +
  font("xy.text", size = 11, color = "black") +
  font("legend.title", size = 14, color = "black") +
  font("legend.text", size = 11, color = "black")
leuk_plot2
ggsave(filename = file.path(images_dir, "leuk_veri.tiff"), width = 10, height = 5, device='tiff')
```

